Higher-Order unification for free!

### 2024-04-15 paper/main.tex
typos

\begin{elpicode}
  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
  con C ~\Ee~fcon C.
  app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
  uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
\end{elpicode}

@@

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of \emph{steps} of length $\mathcal{N}$.
At each step, $p$ we unify two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, which is the result of the
logic program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent
In order to simulate a \Fo{} logic program in \Ho{}, we compile
each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produces a variable mapping $m$ and a list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.
We write $\hoUnifPb_p = \{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.

@@

\paragraph{Backtracking} We omit it from our model of a logic program's execution
since it plays a very minor role, orthogonal to higher-order unification.
We point out that each \emph{run} corresponds to a (proof search) branch in the
logic program that either fails at some point, or succeeds. A computation that
succeeds by backtracking, exploring multiple branches, could be
modeled as a set of runs with (possibly non-empty) common prefixes.

@@

\noindent
In other words the compiler outputs terms in \wellb, even if its
input is not.
Note that the property holds for any substitution. $\rho$ could be given by an
oracle and/or not necessarily be a most general one: in \wellb \Ue{} simply
does not contradict \Eo.

>>>>

\begin{elpicode}
  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
  con C ~\Ee~con C.
  app A ~\Ee~app B :- forall2 ~(\Ee)~ A B.
  lam F ~\Ee~lam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
  uva N A ~\Ee~uva N B :- forall2 ~(\Ee)~ A B.
\end{elpicode}

@@

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of \emph{steps} of length $\mathcal{N}$.
At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, which is the result of the
logic program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent
In order to simulate a \Fo{} logic program in \Ho{}, we compile
each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produces a variable mapping $m$ and a list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.
We write $\foUnifPb_p = \{~ \foUnifPb_{p_l}, \foUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.

@@

\paragraph{Backtracking} We omit it from our model of a logic-program execution
since it plays a very minor role, orthogonal to higher-order unification.
We point out that each \emph{run} corresponds to a (proof search) branch in the
logic program that either fails at some point, or succeeds. A computation that
succeeds by backtracking, exploring multiple branches, could be
modeled as a set of runs with (possibly non-empty) common prefixes.

@@

\noindent
In other words the compiler outputs terms in \wellb, even if its
input is not.
Note that the property holds for any substitution, $\rho$ could be given by an
oracle and/or not necessarily be a most general one: in \wellb \Ue{} simply
does not contradict \Eo.
